---
title: "Final-Project"
format: html
editor: visual
embed-resources: true
---

## Preamble

In this lab, you will explore data from NASA's Exoplanet Archive. This archive contains information about confirmed exoplanets discovered outside our Solar System. This dataset includes information about confirmed exoplanets, such as planet name, host star, discovery method, discovery year, discovery facility, planet radius, planet mass, orbital period, star characteristics, and distance form Earth.

[NASA Exoplanet Archive Data Column Definitions](https://exoplanetarchive.ipac.caltech.edu/docs/API_PS_columns.html)

## Loading Libraries

```{r}

library(tidyverse)
```

## Data

[Planetary Systems Composite Data Download](https://exoplanetarchive.ipac.caltech.edu/cgi-bin/TblView/nph-tblView?app=ExoTbls&config=PSCompPars)

```{r}

PSCD <- read.csv("Planetar_Systems_CompositeData.csv",
                comment.char = "#")
```

**1. How many discovery methods have successfully discovered a new exoplanet? Create a barchart showing the number of planets discovered by each discovery method.**

```{r}

count <- PSCD |>
  count(discoverymethod)

ggplot(count,
       aes(x = reorder(discoverymethod, n),
              y = n)) +
         geom_col() +
         labs(title = "Number of Exoplanets discovered by each discovery method",
              x = "Discovery Method",
              y = "Number of Planets") +
  coord_flip()
         
       


```


**1. Create a function that categorizes each exoplanet into a size. Use classification similar to a Milky Way Galaxy planet.**


```{r}

#Use pl_rade 

size <- function(radius) {
  if(is.na(radius)){
    return("Unknown radius")
  } else if (radius < 1.25) {
    return("Simiar size to Earth")
  } else if(radius < 6){
    return ("Similar size to Neptune")
  } else{
    return("Similar size to Jupiter or larger")
  }
}

PSCD$planet_type <- sapply(PSCD$pl_rade, size)
table(PSCD$planet_type)

```
