---
title: "Final-Project"
format: html
editor: visual
embed-resources: true
---

## Preamble

In this lab, you will explore data from NASA's Exoplanet Archive. This archive contains information about confirmed exoplanets discovered outside our Solar System. This dataset includes information about confirmed exoplanets, such as planet name, host star, discovery method, discovery year, discovery facility, planet radius, planet mass, orbital period, star characteristics, and distance form Earth.

[NASA Exoplanet Archive Data Column Definitions](https://exoplanetarchive.ipac.caltech.edu/docs/API_PS_columns.html)

## Loading Libraries

```{r}

library(tidyverse)
library(knitr)
```

## Data

[Planetary Systems Composite Data Download](https://exoplanetarchive.ipac.caltech.edu/cgi-bin/TblView/nph-tblView?app=ExoTbls&config=PSCompPars)

```{r}

PSCD <- read.csv("Planetar_Systems_CompositeData.csv",
                comment.char = "#")
```

**1. How many discovery methods have successfully discovered a new exoplanet? Create a barchart showing the number of planets discovered by each discovery method.**

```{r}

count <- PSCD |>
  count(discoverymethod)

ggplot(count,
       aes(x = reorder(discoverymethod, n),
              y = n)) +
         geom_col() +
         labs(title = "Number of Exoplanets discovered by each discovery method",
              x = "Discovery Method",
              y = "Number of Planets") +
  coord_flip()
         
       


```

**2. Create a function that categorizes each exoplanet into a size. Use classification similar to a Milky Way Galaxy planet.**

```{r}

#Use pl_rade 

size <- function(radius) {
  if(is.na(radius)){
    return("Unknown radius")
  } else if (radius < 1.25) {
    return("Simiar size to Earth")
  } else if(radius < 6){
    return ("Similar size to Neptune")
  } else{
    return("Similar size to Jupiter or larger")
  }
}

PSCD$planet_type <- sapply(PSCD$pl_rade, size)
table(PSCD$planet_type)

```

**3. How are the different discovery methods related to a planet's proximity from Earth? Make an informative bar chart displaying the difference in the most common distances between each method.**

```{r}
method_profile <- function(method) {
  PSCD |>
    filter(discoverymethod == method) |>
    summarize(
      discoverymethod = method,
      number_of_planets = n(),
      median_distance = median(sy_dist, na.rm = TRUE)
    )
}

method_table <- unique(PSCD$discoverymethod) |>
  map_dfr(method_profile) |>
  filter(number_of_planets >= 10) |>
  arrange(median_distance) |>
  mutate(discoverymethod = factor(discoverymethod, levels = discoverymethod))

ggplot(method_table,
       aes(x = median_distance,
           y = discoverymethod)) +
  geom_col(fill = "gray34") +
  geom_text(aes(label = paste("Planets =", number_of_planets)),
            hjust = -0.05,
            size = 3.25,
            fontface = "italic") +
  expand_limits(x = max(method_table$median_distance) * 1.15) +
  labs(
    title = "Median Distance of Exoplanets by Discovery Method",
    subtitle = "(minimum 10 discovered planets per method)",
    x = "Median Distance from Earth (parsecs)",
    y = "Discovery Method"
  ) +
  theme(axis.ticks.y = element_blank()) +
  theme(axis.title.y = element_text(face = 'bold', margin = margin(r = 15)),
        axis.title.x = element_text(face = 'bold'),
        axis.text.y = element_text(size = 8, margin = margin(r = -15))
)
```

**4. Does the distance (sy_dist) of a planetary system from Earth predict the number of planets (sy_pnum) detected? **

sy_pnum = 1.864 - .0001384(sy_dist)
Due to a p-value of < 2 × 10⁻¹⁶) there is a statistically significant negative relationship between the distance of a planetary system and the number of planets detected. The effect is weak as R^2 = 0.021

```{r}

dist_planets_lm <- lm(sy_pnum ~ sy_dist, 
                    data = PSCD)
summary(dist_planets_lm)

broom::tidy(dist_planets_lm)
broom::glance(dist_planets_lm) |>
  knitr::kable()

ggplot(PSCD, 
       aes(x = sy_dist,
           y = sy_pnum)) +
  geom_point(alpha = .8) +
  geom_smooth(method = "lm") +
  labs(title = "Number of Planets vs Distance from Earth",
       x = "Distance from Earth",
       y = "Number of planets in system") 
  


```